---
title: IKEA, The Flatpack Wizards 🪄
---


```{r}
library(tidymodels, tidyverse)
options(scipen=999)
ikea <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-11-03/ikea.csv')

```


![](https://camo.githubusercontent.com/104f5fbc4bc9adafe09600e6e8fccd29e263757d59659be6f67c695992205963/68747470733a2f2f6669766574686972747965696768742e636f6d2f77702d636f6e74656e742f75706c6f6164732f323031362f31302f61705f3131343333343830383135342e6a70673f773d353735)

IKEA sells much of the furniture found in dorm rooms around the world AND they
happen to sell my favorite chair, the Poäng. They pack it flat, and ship it
around the world. But that comes with cost and their flatpack "technology" is
one way they keep their prices down. Let's explore the wonderful world of IKEA.



# Data

| variable            | description                                                                                         |
|:--------------------|:----------------------------------------------------------------------------------------------------|
| `item_id`           | item id which can be used later to merge with other IKEA dataframes                                 |
| `name`              | the commercial name of items                                                                        |
| `category`          | the furniture category that the item belongs to                                                     |
| `price`             | the current price in Saudi Riyals as shown on 4/20/2020                                             |
| `old_price`         | the price of item in Saudi Riyals before discount sellable_online                                   |
| `sellable_online`   | is the item available online?                                                                       | 
| `link`              | the web link of the item                                                                            |
| `other_colors`      | if other colors are available for the item, or just one color as displayed in the website (Boolean) |
| `short_description` | a brief description of the item                                                                     |
| `designer`          | The name of the designer who designed the item. this is extracted from the full_description column. |
| `depth`             | Depth of the item in Centimeter                                                                     |
| `height`            | Height of the item in Centimeter                                                                    |
| `width`             | Width of the item in Centimeter                                                                     |



# Task 01

- **Question:** How many "Nursery furniture" items are in this data?
- **Answer:** 
- Helpful Hints:
    - Look at your data and think about which columns might give us you an
      answer to this devilishly difficult question.

```{r}




```



# Task 02

- **Question:** How much does the average chair cost at IKEA, in Saudi Riyals? What is the standard deviation of that cost?
- **Answer:** 
    
```{r}


```



# Task 03

- **Question:** What is the cost of the cheapest and most expensive item of
  furniture in the "Sofas & armchairs" category?
- **Answer:** 
- Helpful Hints:
    - You need TWO answers and they should be quite different.

```{r}


```



# Task 04

- **Question:** Does there appear to be a linear relationship between width and
  price for furniture in the "Sofas & armchairs" category? Draw this
  relationship in ggplot and use it to support your answer.
- **Answer:** 

```{r}



```



# Task 05

OK, I'll admit it, the answer to #4 is yes. But if you don't upload your plot,
you don't get any credit for the question. Now I want you to build a linear
regression model of the relationship between the dependent variable, price, and
the independent variable width for furniture in the "Sofas & armchairs"
category.

- **Question:** What is the coefficient of width and what does this coefficient tell us?
- **Answer:** 
- Helpful Hints:
    - There's a gotcha to this one. Our data contains over 3,500 rows of data
      and only SOME of those rows are in the our category of interest.
    - The easiest way to handle this is to build your linear regression against
      a new data set.
    - So create a new data set, `sofas`, which contains our data.

```{r}


```


# Task 06

Get the residuals from your model from the previous task.

- **Question:** What is the average and standard decision of the residuals from
  this model and do they appear to normally distributed? For the second part of
  the question, use your plot to support your answer.
- **Answer:** 
- Helpful Hints:
  - The average residual is going to be a multiple choice. Don't sweat the
    number.

```{r}



```



# Task 07

Develop your own model of price, for sofas, using any feature or combination of
features other than the residuals/predictions of our previous model. Your goal
is to decrease the standard deviation of the residuals and make the distribution
of the residuals more normalized. In Canvas, tell me how you went about
developing an improved model, what features you tried, and what features your
kept. I want to know your formula (y~x) and I want you to prove that your
model's residuals have a smaller standard deviation and are more normally
distributed.

- **Answer** Go to Canvas.

```{r}



```
