---
title: "Covid19 Edition"
output: html_notebook
---


# Goals

- You will work on a real, changing data set.
- You will use the county-grain Covid19 data to assess the status of
  the outbreak here in New York State.
- Honestly, this should connect what we have been doing/learning in
  class and show you how these are genuinely useful skills.
- If this lab doesn't convince you these skills are useful, I give up. :-)



# Setup

- As with the lecture, there's a lot going on in this setup chunk.
    - I have added comments to help you follow what is going on in this code.
    - LMK where it seems confusing.

```{r}
rm(list = ls())
library(ggiraph)
library(knitr)
library(lubridate)
library(maps)
library(rmarkdown)
library(scales)
library(tidyverse)
options(scipen=999)

## There are two data sets, but be aware that the full file, commented out below, is too big for Excel.
data_url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
##data_url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-recent.csv"
which_state <- "New York"
which_county <- "Albany"


county_cases <- 
  read_csv(data_url_counties) %>%
  ## Only keep data for the value of which_state.
  ## This helps everything below run faster.
  filter(state == which_state)

## This is similar to what we did with the State data, but for NY Counties ONLY.
county_data <-
  county_cases %>%
  ## Removes stump rows. Weirdly, New York County was not given a row. Dunno why.
  filter(county != "Bronx", county != "Kings", county != "Richmond", county != "Queens")  %>%
  ## This puts the columns into a nice order.
  select(state,county,fips,date,cases,deaths) %>%
  mutate(
    county       = case_when(county == "New York City"~"Queens",TRUE~county),
    fips         = as.numeric(case_when(county == "Queens"~"36081",TRUE~fips)),
    subregion    = tolower(county),
    daily_cases  = cases - lag(cases, default = first(cases, order_by = county), order_by = county),
    daily_deaths = deaths - lag(deaths, default = first(deaths, order_by = county), order_by = county),
    week_of      = ceiling_date(date, unit="weeks") - 1
    ) %>%
  arrange(state,county,fips,date) %>%
  ## This is a little cruel. It exaggerates the severity of the situation in Queens.
  ## It also creates a new date column called week_of.
  group_by(state,county,fips,subregion) %>%
  mutate(
    daily_cases  = cases - lag(cases, default = first(cases), order_by = county),
    daily_deaths = deaths - lag(deaths, default = first(deaths), order_by = county)
  ) %>%
  ungroup() %>%
  ## These next two steps remove date, and replace it with week_of.
  ## This helps to reduce the noise in the data because of accidents in how
  ## or when data was reported to the state.
  group_by(state,county,fips,subregion,week_of) %>%
  summarize(
    ytd_cases = last(cases),
    ytd_deaths = last(deaths),
    avg_daily_cases  = mean(daily_cases),
    avg_daily_deaths = mean(daily_deaths),
    n_days = n()
  ) %>%
  ## This just prevents us from struggling with it later.
  ungroup() %>%
  ## Have you noticed I am very particular about how my data is arranged?
  select(state,
         county,
         fips,
         subregion,
         week_of,
         n_days,
         ytd_cases,
         avg_daily_cases,
         ytd_deaths,
         avg_daily_deaths) %>%
  ## Only keep 2021 data and only keep complete weeks.
  ## This will remove the incomplete current week.
  ## And it will mean chopping off the first few days.
  filter(week_of > ymd("2020-12-31"), n_days == 7)

## This is a variable that lets us select only the must current, complete week.
current_complete_week <- 
  county_data %>% 
  filter(n_days == 7) %>% 
  summarize(ccw = max(week_of)) %>% 
  pull(ccw)

## This is the population data we will use later.
county_populations <-
    read_csv("https://data.ny.gov/api/views/krt9-ym2k/rows.csv?accessType=DOWNLOAD&sorting=true") %>%
    filter(Year == 2019) %>%
    select(fips = "FIPS Code", population = Population)
big_12 <- (county_populations %>% arrange(desc(population)) %>% pull(fips))[1:12]

## We have replaced the "date" column with "week_of" which 
## is the last day of the week (Saturday).
county_data
```

## Setup Results

That was *a lot* to do in one code chunk. The setup chunk produced:

- county_cases: The county-grain data from the NY Times.
    - Each county in NYS has one row per day.
    - We will not use this data set any further.
- county_data: The county-grain data we created locally.
    - Use this data set for further analysis!!!
    - Each count in NYS has one row per week!
    - Adds population data.
    - Adds additional information such as rolling_new_cases
    - Only has 2021 data for the state in `which_state` which was originally New York.
- county_populations: The population data for each county in NYS, from the New York State government.
    - This analysis uses 2019 estimates because those are the newest numbers I could find from NYS government.
- big_12: A vector of the biggest 12 counties in NYS.
- current_complete_week - A variable giving us the most current complete week.



# Lab Tasks

## Task 01 Add Population Data

- In the setup chunk above, we created a county_data data frame AND a county_populations data frame.
- Join these two data frames by the "fips" column.
    - Use an `inner join()`.
    - Remember, the column you join by must be in quotes `by = "fips"`.
- Similar to our lecture, calculate:
    - year-to-date cases per 100K 
        - Example: `ytd_cases/population*100000`
        - Create a new column called `ytd_cases_100k`
        - The `mutate()` function is your friend.
    - average daily cases per 100K
        - Example: `avg_daily_cases/population*100000`
        - Create a new column called `avg_daily_cases_100k`
- Save all of this to a `county_data`.
    - You already have a data frame called `county_data`.
    - Write some code to produce the requested output and THEN create a new `county_data` with an assignment operator (`<-`).
    - Hint: `county_data <- county_data %>% bunch_of_stuff()`
- If you are feeling stumped, we did something very similar in the lecture notes. . . .

```{r}
## YOUR CODE HERE!!!

```



## Review County Data:

It is important to understand/inspect your data.

- **Data Source:** [NYT Covid19 Data](https://github.com/nytimes/covid-19-data/blob/master/README.md)
- **WARNING:** This analysis assigns all cases/deaths in New York City to Queens.
    - The raw data source assigns all coronavirus cases in the five boroughs to "New York City", which we cannot graph geographically.
    - And, because Queens has more cases than any other borough, this analysis attributes all cases/deaths to that borough/county.
    - We did this above in the setup chunk.
- I suggest you review this data and seek to understand what is here, the grain of the data, and look for any mistakes, etc.
- This is a good time to double-check that you have columns such as avg_daily_cases_100k

```{r echo=FALSE}
county_data
```

We will continue to look at and focus on the growth and distribution of cases throughout New York State.


## Task 02: Plot year-to-date cases by week

- This is similar to what we did in class.
- X Axis: week_of, Y Axis: ytd_cases, color = county
- Only include counties in (`%in%`) the big_12 counties.
    - Filter by fips.
    - Remember, you can filter your data before it gets to the ggplot function.
- Upload this to Canvas.

```{r} 
## YOUR CODE HERE!

```

- Labs rarely call for a moment of silence.
- This is an exception.
- The numbers for Queens are exaggerated because they include cases for the other four boroughs as well.
- And yet, the scale of this tragedy is staggering. 


## Task 03

What is the average number of daily new cases per 100K residents for Albany County for the most recent complete week?

- You are looking for the value of `avg_daily_cases_100k`, which you created earlier.
- You created a variable while running the setup chunk that can help you identify the current full week.

```{r}
## YOUR CODE HERE!
county_data %>% filter(county == "Albany", week_of == current_complete_week) %>% arrange(desc(avg_daily_cases_100k))
```




# Maps!

## Task 04: Draw New York State

You will draw a state map with county lines. We did not do this in class, so I'm going to help you.

- Draw a map which shows the regional distribution of year-to-date cases.
- Review this map and tell me what you think it is telling you in Canvas.
- Because there have been SOOOOOO many cases in NYC (drawn as Queens), it is difficult to tell how our data is distributed otherwise.
- Fortunately, log comes to our aid again! (Have you noticed how useful this function is?)
    - We want to compress our variance so we can get a better view of how the data is distributed.
- Run this map once as I've written it. 
    - Look at the legend, do you see any funky compression in the distribution of the numbers?
- And then run it again, commenting out the line with `scale_fill_gradient()`.
    - This second version is nearly incomprehensible.
    - I'm going to rename this class Intro to the Logarithm.
    - Look at how the numbers shown in the legend change.
    - This second version is what I want you to upload to Canvas.

```{r}

## First, let's load some map data and look at it.
## You only need to load this in once per sussion.
state_map <- map_data("state") %>% filter(region == "new york")
county_map <- map_data("county") %>% filter(region == "new york")

## Now let's build our data set.
## See how we are re-using our max_dates data?
## And . . . . see how this data transformation stuff IS useful?
## When doing REAL statistics.

## This step takes a minute.
## This creates a NEW data set which includes the values we want mapped . . . . 
## and the mapping data to map them!
## And look! This is using an inner join!
county_map_data <- 
  county_data %>%
  filter(week_of == current_complete_week) %>% 
  inner_join(county_map, by = "subregion")

ggplot(county_map_data, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = ytd_cases), color = "white") +
  ## This is the part that magically logs our data.
  scale_fill_gradient(name = "YTD Cases", trans = "log", labels = label_comma()) +
  theme_classic() +
  ggtitle(paste("Distribution of Coronavirus Cases YTD:", current_complete_week))
```


## Task 05: Cases Per 100,000 County Residents

- Take the code from above and alter it so that it shows the distribution of average daily cases per 100,000 county residents.
- You have a column that can do this.
- Be sure to keep the scale_fill_gradient function working, so the data is scaled via the log of avg_daily_cases_100k.
- Look at the distribution of the numbers on the legend.

```{r}
## YOUR CODE HERE!

```
