---
title: "R GIS"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 2
    theme: cerule
---

# Setup

```{r setup, warning = FALSE, message = FALSE}
rm(list = ls())
library(ggiraph)
library(knitr)
library(lubridate)
library(maps)
library(rmarkdown)
library(tidyverse)
options(scipen=999)

data_url_states <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
state_cases <- read_csv(data_url_states)

state_populations <- read_csv("data/state_populations.csv")

## Synthetic data can be useful to explore ideas/relationships.
synthetic_data <- 
  tibble(
    day = 1:10,
    date = seq(ymd('2021-01-01'),ymd('2021-01-10'),by='days'),
    ## Disease A doubles every 24 hours.
    cases_a = c(1,2,4,8,16,32,64,128,256,512),
    log_cases_a = log(cases_a),
    ## Disease B increases by 2 every 24 hours.
    cases_b = 1:10*2,
    log_cases_b = log(cases_b)
  )
```



# Tonight's Goals

- Tell a story with data.
- And this is an important story.
- As such, I will suppress some of the code where I don't think it is important, but we will definitely discuss plenty of code tonight.
- You have all of the code in lecture.Rmd.
- Enjoy!



# Exponential Growth

**Question:** How would you define exponential growth?

Examples of exponential growth:

- compound interest (for good and bad)
- epidemics/pandemics
- fire
- the population of bunnies

## Synthetic Data

- Contains cases data for two imaginary diseases (Disease A and Disease B).
- The MATH we are discussing is applicable beyond epidemiology.
- We have case count records for 10 days.
- One disease is experiencing exponential growth after starting with a single case.
- One is experiencing linear growth after starting with two cases.
- Both diseases start spreading on the first day of 2021.
- One days one - four, the differences do not seem important.
- By day five, the differences become apparent.

```{r echo=FALSE}
synthetic_data %>% paged_table()
```

If real, this would be a terrifyingly FAST moving disease.

- This kind of growth is more akin to bacteria growing in your body.
- It is a synthetic example.
- I wanted the first few days to look similar.


## Exponential Growth Curve 

What does this growth curve look like?

```{r echo=TRUE}
ggplot(synthetic_data, aes(date, cases_a)) +
  geom_point() +
  geom_line() +
  ggtitle("Exponential growth: Do I look familiar?")
```

From Wikipedia:

> It occurs when the instantaneous rate of change, the derivative, of a quantity with respect to time is proportional to the quantity itself.

The number of cases of Disease A doubles every 24 hours ($n^2$). Unfortunately, we have all become far too familiar with exponential growth curves in the evening news.

- But how can we test to see if a relationship is exponential?
- And how can we figure out what the exponent is?

To answer both of these questions, we can exploit the value of the logarithm.

```{r echo=TRUE}
ggplot(synthetic_data, aes(date, log_cases_a)) +
  geom_point() +
  geom_line() +
  ggtitle("May the logarithm set you free")
```

- TADA!
- This should also teach you something about log/exponents.
    - Exponential growth INCREASES DRAMATICALLY as X increases.
    - Log growth DECREASES DRAMATICALLY as X increases.
- Your eyes are actually really good at seeing a straight line. If it looks straight, it is close enough.
    - This example uses fake data. In reality, there would be more noise.

So, that answers the first question. We know how to test for exponential growth, but how do we figure out what the exponent is? Well, that too is pretty easy at this point.

```{r echo=TRUE}
## You've already learned how to calculate a linear model.
## PLEASE NOTE: We are using log_cases_a, not the actual number of cases.
lin_mod <- lm(log_cases_a~date, synthetic_data)
lin_mod
```

- A computer stores dates as an integer.
- What do you suppose the value of `exp(0.693147)` is?
    - If you guessed, `r exp(0.693147)`, congrats!
    - I took the float out a bit further to avoid rounding problems.
- The best math is circular math.



## Not Exponential Growth

If you plot the logs of a linear relationship, like Disease B, you get something else entirely.

Moral of this story:

- Logging your data can be useful.
- But don't just do it, plot the results and let your eyes guide you.
- Because if you do a linear regression in ↓, you get garbage.

```{r echo=TRUE}
ggplot(synthetic_data, aes(date, log_cases_b)) + geom_point() + geom_line()
```



# A Michigan Problem?

<div class="covid-act-now-embed" data-fips-id="26" /><script src="https://covidactnow.org/scripts/embed.js"></script>

You can compare our results to what you can see on [Covid Act Now](https://covidactnow.org/us/michigan-mi/?s=1755676) (above).

- April 13: [Michigan officials request help as COVID-19 cases surge](https://www.cbsnews.com/news/michigan-covid-cases-gretchen-whitmer-help-request/)
- April 13: [Opinion: Vaccines won’t save Michigan from its covid-19 surge](https://www.washingtonpost.com/opinions/2021/04/13/vaccines-wont-save-michigan-its-covid-19-surge/)
- April 16: [Coronavirus: Michigan protesters defy stay-at-home order](https://www.bbc.com/news/av/world-us-canada-52304832)


## Our Data

Tonight, we will tell a story with data.
 
- `state_cases`: From the NY Times: https://github.com/nytimes/covid-19-data
    - You wanted to spend some time with real data, this is real data.
    - And Michigan is a real discussion topic.
- `population`: From Wikipedia, July 2020 population estimates.

**Questions:** 

- Why do we need state population data?
- How do we need to transform this data in order to use discuss the current events in Michigan?

```{r echo=FALSE}
state_cases %>% paged_table()
```

```{r echo=FALSE}
state_populations %>% paged_table()
```


## Data Transformations

- Our analysis depends on a weekly rolling average of new cases/deaths per 100,000 residents.
- The NY Times Covid Data is daily (good), but it is YTD (bad).
- This code groups the covid data by week:
    - calculates rolling new case/death averages for the week
    - and provides data as of the end of the week (total cases/deaths YTD)

```{r echo=TRUE}
## Let's discuss this, but I do not expect you to be able to do this.
state_data <- 
  state_cases %>% 
  select(state, fips,date, cases, deaths) %>%
  arrange(state,fips,date) %>%
  group_by(state,fips) %>%
  mutate(
    new_cases  = cases - lag(cases, default = first(cases), order_by = state),
    new_deaths = deaths - lag(deaths, default = first(deaths), order_by = state)
  ) %>%
  ungroup() %>%
  mutate(week_of = ceiling_date(date, unit="weeks") - 1) %>%
  group_by(state,week_of) %>%
  summarize(
    cases = last(cases),
    deaths = last(deaths),
    new_cases_rolling = mean(new_cases),
    new_deaths_rolling = mean(new_deaths)
  ) %>%
  ungroup() %>%
  left_join(state_populations, by = "state") %>%
  mutate(
    cases_100k = cases/population*100000,
    deaths_100k = deaths/population*100000,
    new_cases_rolling_100k = new_cases_rolling/population*100000,
    new_deaths_rolling_100k = new_deaths_rolling/population*100000,
  )
state_data %>% paged_table()
```

## 2021 Case Counts

- Graph ONLY shows 2021 data.
- You KNOW Michigan is a small state, but LOOK at where they are on that curve.

```{r echo=TRUE}
new_cases <-
  state_data %>% 
  filter(week_of >= "2021-01-01") %>% 
  ggplot(aes(x = week_of, y = new_cases_rolling, color = state)) +
  geom_line_interactive(
    aes(tooltip = state, data_id = state),
    ## Suppressing the legend, because otherwise it (50 state names)
    ## overwhelms everything else.
    show.legend = FALSE
  )
x <- girafe(ggobj = new_cases)
x <- girafe_options(
  x = x,
  opts_hover(css = "stroke:red;stroke-width:2")
)
x
```


## 2021 Case Counts Per 100,000

- I do not understand the Missouri spike in the middle of March. (Data dump?)
- Our results are similar to Covid Act Now, although those are using different data and slightly superior methodology.

```{r}
new_cases_100k <-
  state_data %>% 
  filter(week_of >= "2021-01-01", !state %in% c("Guam","Puerto Rico")) %>% 
  ggplot(aes(x = week_of, y = new_cases_rolling_100k, color = state)) +
  geom_line_interactive(
    aes(tooltip = state, data_id = state),
    show.legend = FALSE
  )
x <- girafe(ggobj = new_cases_100k)
x <- girafe_options(
  x = x,
  opts_hover(css = "stroke:red;stroke-width:2")
)
x
```

```{r}
## As of April 15, Covid Act Now claims 78.7 deaths per 100K and we are
## calculating 76.02, which is pretty decent since we are using different data
## and a slightly less sophisticated methodology.
state_data %>% 
  filter(state == "Michigan") %>% 
  arrange(desc(week_of)) %>% 
  paged_table()
```

- It appears the new case rate may be leveling off, albeit at a rate much higher than anywhere else in the nation.
- Is Rhode Island the next "big" story?
- Let's look at Michigan's numbers in March/April.

```{r echo=TRUE}
michigan_new_cases <-
  state_data %>% 
  filter(week_of >= "2021-03-01", state == "Michigan") %>% 
  ggplot(aes(x = week_of, y = new_cases_rolling_100k)) +
  geom_line_interactive(
    aes(tooltip = state, data_id = state),
    show.legend = FALSE
  ) +
  geom_point_interactive(
    aes(tooltip = new_cases_rolling_100k, data_id = new_cases_rolling_100k), 
    show.legend = FALSE,
    size = 2
  )
  
x <- girafe(ggobj = michigan_new_cases)
x <- girafe_options(
  x = x,
  opts_hover(css = "stroke:red;stroke-width:2")
)
x
```


## How Fast Was It Growing?

- The steepest/scariest part of the slope here is between March 14 and April 4.
- During this time period, how quickly were the number of new cases growing per week?

```{r echo=TRUE}
michigan_model_data <- 
  state_data %>% 
  filter(between(week_of, ymd("2021-03-14"), ymd("2021-04-04")), state == "Michigan") %>%
  mutate(log_new_cases_rolling = log(new_cases_rolling))

michigan_model <- lm(log_new_cases_rolling~week_of, michigan_model_data)
michigan_model
```

```{r}
exp(0.05676)
```


And, if we have time, we should unpack what this means.
