---
title: ACNH Stalk Market!
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

breaks <- hour(hm("00:00", "11:59", "23:59"))
labels <- c("Morning", "Afternoon")

turnips <-
  read_csv("data/turnips.csv") |>
  mutate(
    day = factor(wday(date, label = TRUE)),
    week = case_when(
      year(date) == 2021~epiweek(date)-14,
      date == "2022-01-01"~38,
      year(date) == 2022~epiweek(date)+38,
      year(date) == 2023~epiweek(date)+90,
      TRUE~NA
    ),
    time = hm(time),
    time_of_day = cut(
      x=hour(time),
      breaks = breaks,
      labels = labels,
      include.lowest=TRUE
    )
  ) |>
  select(
    sample = rownames,
    date,
    week,
    day,
    time,
    time_of_day,
    price
  )

```



# Task 01

This data set has 1,419 samples in it. The `week` column assigns every sample
day to a specific week. Every week starts on a Sunday and ends on a
Saturday. There are no incomplete weeks in this data.

- **Question:** How many weeks of data do we have in this sample?
- **Answer:** 110

```{r}

# Students are unlikely to know the nrow function.
# But they should know how to group by week and then check to see how many rows
# of data they have in the result set.
turnips |> count(week) |> nrow()

# Another way to do this:
turnips |> filter(day == "Sun") |> count()

```



# Task 02

Draw a boxplot of sale price for each day of the week, except Sunday.

- **Question:** Based on this plot, which days (on average) seem to be the best
  days to sell your turnips on and which days would you avoid?
- **Answer:** Wednesday/Thursday look like the best days. Monday/Saturday look
  like the worst days, on average.
        - To get credit, they don't need to match my answer, but I want them to
          get at least part of it and I want them to explain their logic.

```{r}

turnips |>
  filter(day != "Sun") |>
  ggplot(aes(x = day, y = price)) +
  geom_boxplot()

# Take off a point or two if they include Sunday. B/C you can't seel on Sunday.

```



# Task 03

- **Question:** What is the average sale price of Daisy Mae's turnips?
- **Answer:** 100

```{r}

turnips |>
  filter(day == "Sun") |>
  summarize(avg_price = mean(price, na.rm = TRUE))

```



# Task 04

Let's draw a new kind of plot, a trend line. You can do this by defining the x
aesthetic to the date and the y aesthetic to price. Use geom_line to, well, draw
the line. Use this plot to get some new ideas for the best day/time to sell
turnips.

- **Question:** Work as a team to develop a strategy to sell turnips which would
maximize your profits. Using your box plot, line plot, and any other plots or
information you make have created, tell me how you would maximize your turnip
sale price. Don't spend more than 15 minutes on this, but do spend some time
trying to maximize your 🔔. Be sure to explain what your group looked for and
your process for determining the best sales strategy.
- **Answer:** This is hard to grade, see below, I'm looking for their strategy
  as much as their answers.

```{r}

turnips |>
  filter(day != "Sun") |>
  ggplot( aes(x = date, y = price)) +
  geom_line()

# This one is a little hard to grade.
# IF they use just these two plots, give them 80% of the credit.
# If they try to figure out when the spikes occur, given them 100%
# I'm looking for some evidence they tried to develop a way to ID times/days
# where the price of a turnip is over 100.
# Or that they went to find a pattern for identifying when the spikes would
# occur.

turnips |> filter(price > 100) |> count(time_of_day)

# and/or

turnips |> filter(price > 300) |> count(day)

```



# Task 05

Although this data is pretty good, we are missing some data.

- **Question:** How many rows of data are we missing and which days have the
  most missing data?
- **Answer:** There are six missing price samples. Monday and Thursday are the
  "worst" with two missing each.

```{r}

turnips |> filter(is.na(price)) |> count(day)

```

